---
title: "Fast package installation with Bioconductor package binaries - Bioc2021"
author: "Nitesh Turaga"
date: "8/4/2021"
output: 
  html_document:
    theme: sandstone
    highlight: tango
    toc: true
---

***

# Launch bioconductor_docker image

The Bioconductor docker images are available on Dockerhub and the
source code on Github at
https://github.com/Bioconductor/bioconductor_docker

Clone the image,

```{bash, eval=FALSE}
git clone https://github.com/Bioconductor/bioconductor_docker

cd bioconductor_docker
```

Start the image,

```{bash, eval=FALSE}
docker-compose up -d
```

To stop the image later

```{bash, eval=FALSE}
docker-compose stop
```

***

# Demo of traditional installation of Bioconductor packages

Let us first take a look at the repositories from which BiocManager
downloads the tarballs.

```{r biocmanager_repos, message=FALSE, warning=FALSE}
BiocManager::repositories()
```

As you can see they are coming the location https://bioconductor.org/packages/3.13/bioc 
for software packages.

So when we go ahead an install a package by running the command below, two things happen

1. It downloads the tarball for the package from Bioconductor.
2. Then compiles on your local machine specific to your machine (or docker image)

```{r, eval=FALSE}
BiocManager::install('BiocGenerics')
```

The result is shown below is truncated for clarity. Notice the line starting with 
`trying URL`, where the download happens, and then the line
`byte-compile and prepare package for lazy loading` which starts the compilation.

```
'getOption("repos")' replaces Bioconductor standard repositories, see '?repositories' for
details

replacement repositories:
    CRAN: https://packagemanager.rstudio.com/all/__linux__/focal/latest

Bioconductor version 3.14 (BiocManager 1.30.16), R 4.1.0 (2021-05-18)
Installing package(s) 'BiocGenerics'
trying URL 'https://bioconductor.org/packages/3.14/bioc/src/contrib/BiocGenerics_0.39.1.tar.gz'
Content type 'application/x-gzip' length 45497 bytes (44 KB)
==================================================
downloaded 44 KB

* installing *source* package ‘BiocGenerics’ ...
** using staged installation
** R
** inst
** byte-compile and prepare package for lazy loading
Creating a new generic function for ‘append’ in package ‘BiocGenerics’
Creating a new generic function for ‘as.data.frame’ in package ‘BiocGenerics’
<truncated>
Creating a new generic function for ‘dirname’ in package ‘BiocGenerics’
** help
*** installing help indices
** building package indices
** testing if installed package can be loaded from temporary location
** testing if installed package can be loaded from final location
** testing if installed package keeps a record of temporary installation path
* DONE (BiocGenerics)

The downloaded source packages are in
	‘/tmp/Rtmpm2LoxR/downloaded_packages’
```

Now when we take a look at the `.libPaths()` location here, it shows the path
where your packages are being installed.

```{r, echo=TRUE}
.libPaths()
```

Remember that the path `/usr/local/lib/R/host-site-library` is mapped
to your local machine at `$HOME/R/bioconductor_docker/3.14`.

***

# Demo of Binary installation of Bioconductor packages

Binary installations are fundamentally different from traditional
installation. These are pre-compiled on the docker images with all the
system dependencies that are available on it. Let's take a look at the 
repositories from where the binaries are downloaded.

NOTE: This works only on the docker images based on
`bioconductor_docker`.

```{r message=FALSE, warning=FALSE}
AnVIL::repositories()
```

Once the binary package is downloaded on to your machine, 
all you need to do is point your `.libPaths()` to that location.

**The controlled environment on these Bioconductor docker images 
where we know the operating system and system dependencies, 
make the binary packages possible.**

So, when we install a binary package,

1. Download the binary tarball from the cloud,
  
2. Point the `.libPaths()` location to recognize the package.

```{r echo=TRUE, eval=FALSE}
AnVIL::install('rhdf5')
```

Output of the install function shows us that the download of the package is 
taking place from a location at https://storage.googleapis.com. This is an
object store on the Google cloud. Comparing with the previous traditional 
installation, there is no "byte compile" step.

```
'getOption("repos")' replaces Bioconductor standard repositories, see '?repositories' for details

replacement repositories:
    CRAN: https://packagemanager.rstudio.com/all/__linux__/focal/latest

trying URL 'https://storage.googleapis.com/bioconductor_docker/packages/3.14/bioc/src/contrib/rhdf5_2.37.0_R_x86_64-pc-linux-gnu.tar.gz'
Content type 'application/x-tar' length 7234933 bytes (6.9 MB)
==================================================
downloaded 6.9 MB

* installing *binary* package ‘rhdf5’ ...
* DONE (rhdf5)

The downloaded source packages are in
	‘/tmp/Rtmpm2LoxR/downloaded_packages’
```

To get more familiar with the binary installation process, let's take a look 
at the function signature for `AnVI::install()`.

```{r anvil_sig, message=FALSE}
args(AnVIL::install)
```

We can specify the option `binary_base_url` for either Microsoft Azure or
Google cloud (currently Google cloud is the default).

```{r, eval=FALSE, echo=TRUE}
azure <- 'https://bioconductordocker.blob.core.windows.net/packages'
google <- 'https://storage.googleapis.com/bioconductor_docker/packages'

AnVIL::install('Rhtslib') ## Default from Google

## To explicitly install from google
AnVIL::install('Rhtslib', force = TRUE, binary_base_url = google) 

## from Microsoft azure
AnVIL::install('Rhtslib', force = TRUE, binary_base_url = azure) 
```

***

# Benchmark analysis - Binary vs Traditional installation

Let's compare the speed of the two methods of installations. We shall
use the package `microbenchmark` for this.

```{r, eval=FALSE}
## Let's leverage the AnVIL installation to speed this up ;)
AnVIL::install('microbenchmark')
```

Once it's installed, let's compare binary vs traditional installation.

```{r, warning=FALSE, message=FALSE}
res <- microbenchmark::microbenchmark(
    "binary" = {AnVIL::install(c('Rhtslib', 'rhdf5'))},
    "traditional" = {BiocManager::install(c('Rhtslib', 'rhdf5'), force = TRUE)},
    times = 2L ## Run the same process two times
)
```

Print the result in `seconds` as the unit.

```{r echo=TRUE}
print(res, units = 's')
```

We can see that upon running this twice, measured in seconds there is
a clear speed up comparing binary vs traditional installation.

***

# Conclusion - Session Info

```{r}
sessionInfo()
```

